---
title: "Estimation of finite population proportions for small areas – a statistical data integration approach"
author: "Aditi Sen"
date: '2024-02-25'
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
getwd()
```

# Load packages required

```{r setup, include=FALSE}
library(usmap); library(lme4); library(ggplot2); 
library(dplyr); library(optimParallel); library(foreach);library(doParallel);
library(tidyverse); library(maps); library(fastDummies); library(PNWColors); library(purrr); library(survey); library(maps); library(fastDummies);
library(gganimate); library(ggeasy); library(ggpubr); library(patchwork); 
```

# Setup parameters like controls for optimParallel to run

```{r }
if(tolower(.Platform$OS.type) != "windows"){
  cl <- makeCluster(spec=detectCores(), type="FORK", outfile="")  
} else
  cl <- makeCluster(spec=detectCores(), outfile="")
setDefaultCluster(cl=cl)

options("optimParallel.forward", "optimParallel.loginfo")
control01 <- list(factr=.01/.Machine$double.eps)
control001 <- list(factr=.001/.Machine$double.eps)
control0001 <- list(factr=.0001/.Machine$double.eps)
```

## Data reading and preparation for PEW, CPS 

```{r }
data <- readRDS("data.RDS") # read PEW, CPS and actual election data saved in data.RDS
pew_1 <- data[[1]]; cps_1 <- data[[2]]; actual_result <- data[[3]] 

# function for data preparation: takes dataset as input and creates auxilliary variables like age, race, education for modelling

data_prep <- function(data_name){
  data_name$ID <- 1:nrow(data_name)
  data_name$age_new <- ifelse(data_name$age4=="45-64",1,0) # indicator for age group 45-64
  data_name$age_new <- factor(data_name$age_new, ordered = FALSE) # convert into a factor
  dummies <-dummy_cols(data_name,select_columns = 'educ4') # create dummy variables for education categories
  dummies <- dummies %>% rename("educ4_col_grad" = "educ4_col grad")
  old <- data_name; data_name <- dummies
  data_name$educ4_col_grad <- factor(data_name$educ4_col_grad, ordered = FALSE) # indicator for education college grad
  data_name$educ4_postgrad <- factor(data_name$educ4_postgrad, ordered = FALSE) # indicator for education post grad
  return(data_name)
}

pew <- data_prep(pew_1); cps <- data_prep(cps_1)

```

# Using glmer function in lme4 package, run mixed effects model and save coefficient estimates

```{r }

logis_fit <- glmer(formula = demvote ~ age_new + gender + race3 + educ4_col_grad + educ4_postgrad + qlogis(obama12) + (1 | state),
    data = pew, family = "binomial")
summary(logis_fit)
intercept <- fixef(logis_fit)[1] # intercept
age1_coeff <- fixef(logis_fit)[2] # age group 45-64
gen_fem_coeff <- fixef(logis_fit)[3] # gender female
race_blk_coeff <- fixef(logis_fit)[4] # race black
race_hisp_coeff <- fixef(logis_fit)[5] # race hispanic
edu_colgrad_coeff <- fixef(logis_fit)[6] # education college grad
edu_postgrad_coeff <- fixef(logis_fit)[7] # education post grad
obama12coeff <- fixef(logis_fit)[8] # voting percentage for obama
ran <- as.data.frame(VarCorr(logis_fit)); # random effect 
ran$sdcor

```

# Calculate x'beta 

```{r,echo=TRUE}

# function takes input dataset and parameter estimates obtained from modelling

x_beta_func <- function(data_name,bhat0,bhat1,bhat2,bhat3,bhat4,bhat5,bhat6,bhat7) {
  age_cont <- gen_cont <- race_cont <- edu_cont <- obama_cont <- NULL
# the covariates for model : age, gender, race, education (at unit level), obama % (at area level)
  age4 <- data_name$age4; gender <- data_name$gender; race3 <- data_name$race3; educ4 <- data_name$educ4 
  obama12 <- data_name$obama12
  
  for (i in 1:nrow(data_name))
  { # if age variable takes value 45-64 then take beta_1 otherwise 0
    if(age4[i] == "45-64") { 
      age_cont[i] <- bhat1 
    } else { 
        age_cont[i] <- 0
    }
     # if gender variable takes value female then take beta_2 otherwise 0
    if(gender[i] == "female") { 
      gen_cont[i] <- bhat2 
      } else { 
          gen_cont[i] <- 0
      } 
     # if race variable takes value black then take beta_3,  if hispanic then take beta_4 otherwise 0
    if(race3[i] == "black") { 
          race_cont[i] <- bhat3 
          } else if(race3[i] == "hispanic") { 
            race_cont[i] <- bhat4 
            } else {
            race_cont[i] <- 0
            }
    # if education variable takes value college grad then take beta_5,  if post grad then take beta_6 otherwise 0
    if(educ4[i] == "col grad") { 
      edu_cont[i] <- bhat5 
      } else if(educ4[i] == "postgrad") { 
          edu_cont[i] <- bhat6 
          } else { 
              edu_cont[i] <- 0 
                }   
  }
  # multiply obama % x value with beta_7
  obama_cont <- qlogis(obama12)*bhat7
  
  # add all the above with intercept to get final x'beta
  x_beta <- intercept + age_cont + gen_cont + race_cont + edu_cont + obama_cont
  
  return(x_beta)
}

# use above function to get x'beta for PEW and CPS
pew$x_beta <- x_beta_func(data_name=pew, bhat0=intercept, bhat1=age1_coeff, bhat2=gen_fem_coeff, bhat3=race_blk_coeff, bhat4=race_hisp_coeff, bhat5=edu_colgrad_coeff, bhat6=edu_postgrad_coeff, bhat7=obama12coeff)

cps$x_beta <- x_beta_func(data_name=cps, bhat0=intercept, bhat1=age1_coeff, bhat2=gen_fem_coeff, bhat3=race_blk_coeff, bhat4=race_hisp_coeff, bhat5=edu_colgrad_coeff, bhat6=edu_postgrad_coeff, bhat7=obama12coeff)

```

# Prediction using predict function on glmer model object

```{r, echo = TRUE}

# list of states (pew has 49 and cps has 51 states)
pew_state_list <- sort(unique(pew$state))
cps_state_list <- sort(unique(cps$state))

# subset of Montana and SD with only required columns for prediction
cps_sub <- cps %>% dplyr::select(ID,state,x_beta,age_new, gender, race3, educ4_col_grad, educ4_postgrad, obama12, weight)
cps_2 <- subset(cps_sub, subset = state %in% c("MT","SD"))

# subset of 49 other states  with only required columns for prediction
cps_49 <- subset(cps_sub, subset = state %in% pew_state_list)

# calculate the estimated prob from predict function with random effect
cps_49$pred_prob <- predict(logis_fit, newdata = cps_49, re.form = ~(1|state),
                            type = "response")
  
# without random effect for SD and MT
cps_2$pred_prob <- predict(logis_fit, newdata = cps_2, re.form = NA, 
                    type = "response")

# combine all states
cps_all <- as.data.frame(rbind(cps_49, cps_2))
cps_all$syn_prob <- predict(logis_fit, newdata = cps_all, re.form = NA, 
                    type = "response")
length(unique(cps_all$state))

# list of states
state_list <- unique(cps_all$state)
pred_func_results <-  NULL

for(state_name in state_list)
{
  data_state <- state_est <-  NULL
  
  # subset cps data by state
  data_state <- subset(cps_all, state == state_name, 
                       select = c(state, pred_prob, syn_prob, weight))
  
  # calculate weighted sum
  state_est <- c(state_name, 
                 100*sum(data_state$pred_prob*data_state$weight)/
                         sum(data_state$weight),
                 100*sum(data_state$syn_prob*data_state$weight)/
                         sum(data_state$weight)
                 )
  pred_func_results <- rbind(pred_func_results,state_est)
}

pred_func_results <- as.data.frame(pred_func_results)
colnames(pred_func_results) <- c("state","pred_est","syn_est")
rownames(pred_func_results) <- seq(1:51)
```


## User defined function which computes parameter estimates using EM algorithm
## This can be used on simulated datasets with the required columns names

```{r }
# function takes dataset and initial values as inputs parameters
EM_func <- function(input_data,beta_ini,sigma_ini)
{
  # set values of other parameters like number of simulations and bounds
  case_sim=1000;v_lower=-3; v_upper=3;sigma_lower=0;sigma_upper=100;
  # keep only required variables from datasets to save space
  input_data <- input_data %>% dplyr::select(state, age4, gender, race3, educ4,obama12, demvote)
  # list of states
  state_list <- unique(input_data$state)
  # create covariates
  input_data$y <- input_data$demvote
  input_data$x1 <- ifelse(input_data$age4=="45-64",1,0) #age_new
  input_data$x2 <- ifelse(input_data$gender=="female",1,0) #gen_fem
  input_data$x3 <- ifelse(input_data$race3=="black",1,0) #race3black
  input_data$x4 <- ifelse(input_data$race3=="hispanic",1,0) #race3hispanic
  input_data$x5 <- ifelse(input_data$educ4=="col grad",1,0) #educ4_col_grad
  input_data$x6 <- ifelse(input_data$educ4=="postgrad",1,0) #educ4_postgrad
  input_data$x7 <- qlogis(input_data$obama12) #obama12
  # drop variables not needed anymore
  input_data <- subset(input_data, select = -c(age4, gender, race3, educ4,
                                               obama12, demvote))
  # split components of initial parameter vector passed into beta0,..,beta7
  for(i in 1:length(beta_ini)){
    nam  <- paste0("beta",i-1); assign(nam, beta_ini[i])
  }
  
# Step 1 : create x_beta with initial beta values 
  
    input_data$x_beta <- beta0 + input_data$x1*beta1 + input_data$x2*beta2 +
    input_data$x3*beta3 + input_data$x4*beta4 + input_data$x5*beta5 + 
    input_data$x6*beta6 + input_data$x7*beta7 
  
# Step 2 : Maximize f(vi) for area i
    
  # function for defining f(v_i) 
  vhat_func <- function(state_name) {
    g_vi <- function(v_i){
      t <- ((exp(xi_beta + v_i)/(1 + exp(xi_beta + v_i)))^yi)*
        ((1-exp(xi_beta + v_i)/(1 + exp(xi_beta + v_i)))^(1-yi))
      return(exp(-(v_i^2)/(2*(sigma_ini^2)))*prod(t))
    }
    # filter data for i-th state
    data_name <- filter(input_data,state==state_name)
    xi_beta <- data_name$x_beta; yi <- data_name$y; ni <- nrow(data_name)
    # optimization step with bounds on v_i
    opt <- optimize(f = g_vi,lower = v_lower,upper = v_upper,maximum = TRUE)
    # calculate tau_hat_sq with v_i_hat obtained
    tau_hat_sq <- (1/(sigma_ini^2)+
                     sum(exp(xi_beta + opt$maximum)/
                           ((1+exp(xi_beta + opt$maximum))^2)))^(-1)
    # combined dataset with vi hat and tau i from all areas
    r <- cbind(as.character(state_name), ni, opt$maximum, tau_hat_sq)
    res <- rbind(data.frame(), r)
  }
  # run function here on the full dataset
  lap_approx <- data.frame(t(sapply(lapply(state_list, vhat_func),c)))
  # the final dataset contains state, sample size, v_hat, tau_hat_square
  colnames(lap_approx) <- c("state","ni","v_hat","tau_hat_sq") 
  
# Step 3 : draw for ith area an observation from N (v_i_hat,tau_i_hat_sq) and repeat R = 100 times to get v_i tilde
  
  # This function draws random sample from normal and outputs dataset with vi_tilde   
  vtilde_func <- function(trial) {
    v_tilde <- as.numeric(lap_approx$v_hat) + 
      sqrt(as.numeric(lap_approx$tau_hat_sq))*
      rnorm(n=length(unique(state_list)),mean = 0, sd = 1)
    r <- cbind(as.character(lap_approx$state),as.numeric(v_tilde))
    # to return multiple objects
    out <- list()
    out$data_set <- rbind(data.frame(), r)
    out$n_v_sq <- sum((v_tilde^2)*as.numeric(lap_approx$ni))
    return(out)
  }
  vtilde_out <- lapply(1:case_sim, vtilde_func)
  # extract objects from list
  n_v_sq <- lapply(vtilde_out, '[[',2)
  v_case_sim <- lapply(vtilde_out, '[[',1)
  v_case_sim <- lapply(v_case_sim, setNames, c("state","v_tilde"))
  
  # Step 4: Write Q function to maximize
  
  # function to maximize for sigma
  Q_theta_sigma <- function(sigma_sq){
    return(log(sigma_sq)-(nrow(input_data)/2)*log(sigma_sq)-
             mean(unlist(n_v_sq))/(2*sigma_sq))
  }
  Q_theta_sigma <- Vectorize(Q_theta_sigma)
  #curve(Q_theta_sigma, xlim = c(0,10000))
  # optimize here
  opt_sigma <- optimize(f = Q_theta_sigma, 
                        lower = sigma_lower, upper = sigma_upper, 
                        tol = 0.0001, maximum = TRUE)

  # function to optimize for beta
  h2_beta <- function(beta){
    # components of beta
    for(i in 1:length(beta_ini)){
      nam  <- paste0("b",i-1); assign(nam, beta[i])
    }
    NewVar <- function(r) { 
      vr <- v_case_sim[[r]]
      newvar2 <- function(state_name){
        data_name <- input_data[input_data$state==state_name,]
        vri <- as.numeric(vr[vr$state==state_name,]$v_tilde)
        # (CHANGE HERE)
        xb <- b0 + data_name$x1*b1 + data_name$x2*b2 + data_name$x3*b3 +
          data_name$x4*b4 + data_name$x5*b5 + data_name$x6*b6 + 
          data_name$x7*b7
        ti <- -data_name$y*log(exp(xb + vri)/(1 + exp(xb + vri)))+
          (1-data_name$y)*log(1 + exp(xb + vri))
        return(sum(ti,na.rm = TRUE))
      }
      return(sum(sapply(state_list,newvar2)))
    }
    tot <- sapply(1:case_sim,NewVar)
    return(mean(tot))
  }
  # optimize here using optimParallel
  opt_beta <- optimParallel(par=beta_ini, fn = h2_beta,control=control001)
  # The output of this EM function is a list containing beta_hat and sigma_hat and vi sample values
  EM_out <- list()
  EM_out$beta_hat <- opt_beta$par
  EM_out$sigma_hat <- round(sqrt(opt_sigma$maximum),4)
  EM_out$v_case_sim <- v_case_sim
  return(EM_out)
}

```


## Run algorithm on PEW data to get parameter estimates

```{r }
sigma <- 0.1 # initial sigma
for(i in 1:8){ nam  <- paste0("beta",i-1); assign(nam, 0.1)} # initial beta values
# note the start time of the algorithm
EM_start_time <- Sys.time()
# run algorithm iteratively here until convergence i.e. absolute difference between iterations < 0.01
for(k in 1:1000){
  sigmaOld <- sigma 
  for(j in 1:8){
    nam1 <- paste0("beta",j-1,"old"); nam2 <- paste0("beta",j-1)
    assign(nam1,get(nam2))
  }
    ## update
    f_value <- EM_func(input_data = pew,
                       sigma_ini=sigma,
                       beta_ini=c(beta0,beta1,beta2,
                                  beta3,beta4,beta5,
                                  beta6,beta7))
    # get sigma
    sigma <- round(f_value$sigma_hat,4)
    # get beta
    b <- round(f_value$beta_hat,4)
    # assign updated values
    for(j in 1:8){
      nam  <- paste0("beta",j-1); assign(nam, as.numeric(b[j]))
    }
    # condition for convergence
    if(abs(sigma - sigmaOld) < 0.01 & abs(beta0 - beta0old) < 0.01 & 
       abs(beta1 - beta1old) < 0.01 & abs(beta2 - beta2old) < 0.01 &
       abs(beta3 - beta3old) < 0.01 & abs(beta4 - beta4old) < 0.01 &
       abs(beta5 - beta5old) < 0.01 & abs(beta6 - beta6old) < 0.01 &
       abs(beta7 - beta7old) < 0.01) {
        EM_end_time <- Sys.time()
        print(EM_end_time - EM_start_time)
        break
    }
  }
  # function ends here

# print final values 
print(paste0("EM algorithm total iterations:",k))
print(paste0("EM estimates: ","Sigma: ",sigma,
               ", Beta0: ",beta0,", Beta1: ",beta1,
               ", Beta2: ",beta2,", Beta3: ",beta3,
               ", Beta4: ",beta4,", Beta5: ",beta5,
               ", Beta6: ",beta6,", Beta7: ",beta7)) 

```

## Calculate EBP at state level

```{r }
# calculate x'beta with new EM parameter estimates
cps$x_beta_EM <- x_beta_func(data_name = cps, 
                           bhat0=beta0, bhat1=beta1, bhat2=beta2, bhat3=beta3, bhat4=beta4, bhat5=beta5, bhat6=beta6, bhat7=beta7)
  
# calculate probability and generate y
# cps$yij_ebp <- exp(x_beta_EM)/(1+exp(x_beta_EM))
case_sim = 100; v_case_sim <- f_value$v_case_sim  
# function to calculate yij ebp with vi sample values
f1 <- function(state_name){
    data_name <- cps[cps$state==state_name,]
    xb <- data_name$x_beta_EM
    f2 <- function(r){
      vr <- v_case_sim[[r]]
      vri <- as.numeric(vr[vr$state==state_name,]$v_tilde)
      ti <- exp(xb + vri)/(1 + exp(xb + vri))
    }
    chk <- apply(sapply(1:case_sim,f2),1,mean)
    r <- cbind(as.character(state_name),chk,data_name$weight)
    res <- rbind(data.frame(), r)
  }

chk_yij_ebp <- do.call(rbind, lapply(pew_state_list, f1))
colnames(chk_yij_ebp) <- c("state","yij_ebp","weight")
# For MT, SD there is no vi, so do these separately with vi = 0
cps_MTSD <- cps[cps$state %in% c("MT","SD"),c("state","x_beta_EM","weight")]
cps_MTSD$yij_ebp <- exp(cps_MTSD$x_beta_EM)/(1 + exp(cps_MTSD$x_beta_EM))
chk_yij_ebp2 <- rbind(chk_yij_ebp,cps_MTSD[,c("state","yij_ebp","weight")])
    
area_level_stat <- function(state_name,data_name)
  {
    # subset data_b by state
    data_state <- filter(data_name,state==state_name)
    # calculate weighted sum
    state_est <- round(100*sum(as.numeric(data_state$yij_ebp)*
                                 as.numeric(data_state$weight)) / 
                         sum(as.numeric(data_state$weight)),4)
    all_state_est <- cbind(as.character(state_name),state_est)
    res <- rbind(data.frame(), all_state_est)
    return(res)
  }
  
# apply above function on CPS to get EBP Yi bar using new EM parameter estimates
EM_EBP_est <- bind_rows(lapply(cps_state_list, area_level_stat, data_name=chk_yij_ebp2))
colnames(EM_EBP_est) <- c("state","EM_est")

```

# Calculate direct estimates from PEW data and compare with actuals

```{r }
d_design <- svydesign(id=~ID, weights=~weight, data=pew)
direct_est <- as.data.frame(svyby(~demvote,~state,design=d_design, 
                                   FUN = svymean, keep.names = FALSE, 
                                   na.rm = TRUE, covmat = TRUE))
direct_est$direct <- 100*direct_est$demvote
direct_est <- select(direct_est,-c("demvote","se"))

df_list <- list(actual_result, direct_est, pred_func_results, EM_EBP_est)

#merge all data frames in list
comp_data <- df_list %>% reduce(full_join, by='state')
comp_data_49 <- subset(comp_data, subset = state %in% pew_state_list)
# convert columns to numeric
comp_data <- comp_data %>% mutate_at(c('actual', 'direct','EM_est'), as.numeric)

write.csv(comp_data,"comp_data.csv")

```


# Summary measures : ASD, RASD, AAD

```{r,echo=TRUE}

# Average Squared Deviation (ASD)
asd <- function(est, act) (mean((est - act)^2, na.rm = TRUE))
#  EBP estimate vs actual
print(paste0("ASD EBP Est vs Act is ", 
             round(asd(comp_data$EM_est, comp_data$actual),2)))
# direct vs actual
print(paste0("ASD Direct Est vs Act is ",
             round(asd(comp_data_49$direct, comp_data_49$actual),2)))


# square root of ASD (i.e., RASD)
rasd <- function(est, act) sqrt(mean((est - act)^2, na.rm = TRUE))
#  estimate vs actual
print(paste0("RASD EBP Est vs Act is ", 
             round(rasd(comp_data$EM_est, comp_data$actual),2)))
# direct vs actual
print(paste0("RASD Direct Est vs Act is ",
             round(rasd(comp_data_49$direct, comp_data_49$actual),2)))



# Average Absolute Deviation (AAD), where you replace the squared deviation by absolute deviation
aad <- function(est, act) mean(abs(est - act))
#  estimate vs actual
print(paste0("AAD EBP Est vs Act is ", 
             round(aad(comp_data$EM_est, comp_data$actual),2)))
# direct vs actual
print(paste0("AAD Direct Est vs Act is ",
             round(aad(comp_data_49$direct, comp_data_49$actual),2)))

```
# Create US maps

```{r }
pal <- pnw_palette("Bay", n=100, type="continuous")

# Map of actual
plot_act <- plot_usmap(data = comp_data, values = "actual", labels=FALSE) + scale_fill_gradientn(colours = pal, name = "% voters", limits = c(0,96)) + easy_move_legend(to = c("right")) + labs(title = "Actual % voters for Clinton in 2016 election") + theme(panel.background = element_rect(colour = "black")) + theme(plot.title = element_text(hjust = 0.5))
plot_act
# save the plot
#ggsave("actual_usa.png",plot = last_plot(), bg = "white")

# Map of direct estimates
plot_direct <- plot_usmap(data = comp_data, values = "direct", labels=FALSE, include = pew_state_list, show.legend = FALSE) + scale_fill_gradientn(colours = pal, limits = c(0,96)) + labs(title = "Direct Estimate") + theme(panel.background = element_rect(colour = "black")) + theme(plot.title = element_text(hjust = 0.5))

# Map of EBP estimates
plot_EBP <- plot_usmap(data = comp_data, values = "EM_est", labels=FALSE, show.legend = FALSE) + scale_fill_gradientn(colours = pal, limits = c(0,96)) + labs(title = "EBP") + theme(panel.background = element_rect(colour = "black")) + theme(plot.title = element_text(hjust = 0.5))

figure <- ggarrange(plot_direct, plot_EBP,
#                    labels = c("Direct Estimate", "EBP"),
                    ncol = 2, nrow = 1)
figure
# save the plot
#ggsave("direct_EBP_usa.png",plot = last_plot(), bg = "white")

```


#==========================================================================================

## Write bootstrap function for MSPE

```{r }

boot_func <- function(bt){
  print(paste0("start of bootstrap ",bt))
  
  data_pew_b <-data_cps_b<- NULL
  
  ## Step 1 : Subroutine to generate bootstrap data from any dataset
  # function which generates y for each state
  state_func <- function(state_name,data_name) {
    data_state <- data_bi <- demvote <- theta_bi <- v_bi <- NULL
    # subset data by state
    data_state <- filter(data_name,state==state_name)
    # draw from N(0, sigma_hat_sq) where sigma_hat_sq is from model summary
    v_bi <- rnorm(n = 1, mean = 0, sd = ran$sdcor)
    # define prob theta
    theta_bi <- exp(data_state$x_beta + v_bi)/
        (1 + exp(data_state$x_beta + v_bi))
    # draw an observation from Bern with probability theta
    bern_samp <- function(prob){
      rbinom(n = 1, size = 1, prob)
    }
    demvote <- sapply(theta_bi, bern_samp)
    # create combined data for state i
    data_bi <- cbind(demvote,data_state[,c("state","age4","gender","race3","educ4","obama12","weight")])
    # return bootstrap data for state i
    res <- rbind(data.frame(), data_bi)
    return(res)
  }
  
  ## Step 2 : Generate PEW bootstrap sample using above function
  data_pew_b <- bind_rows(lapply(pew_state_list, state_func, data_name=pew))
  print(sum(data_pew_b$demvote))
  
  # run glmer function on PEW Bootstrap sample (not for analysis but for comparison of estimates)
  data_pew_b$x1 <- ifelse(data_pew_b$age4=="45-64",1,0) #age_new
  data_pew_b$x2 <- ifelse(data_pew_b$gender=="female",1,0) #gen_fem
  data_pew_b$x3 <- ifelse(data_pew_b$race3=="black",1,0) #race3black
  data_pew_b$x4 <- ifelse(data_pew_b$race3=="hispanic",1,0) #race3hispanic
  data_pew_b$x5 <- ifelse(data_pew_b$educ4=="col grad",1,0) #educ4_col_grad
  data_pew_b$x6 <- ifelse(data_pew_b$educ4=="postgrad",1,0) #educ4_postgrad
  data_pew_b$x7 <- qlogis(data_pew_b$obama12) #obama12
  
  logis_fit_boot <- glmer(formula = demvote ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + (1 | state), data = data_pew_b, family = "binomial") 
  print(paste0("GLMER SD: ",
               round(attributes(VarCorr(logis_fit_boot)$state)$stddev,4)))
  print("GLMER beta: ")
  print(round(fixef(logis_fit_boot),4))

  ## Step 3 : Run function for EM algorithm on PEW Bootstrap sample to get parameter estimates
  
  sigma <- 0.1 # initial sigma
  for(i in 1:8){ nam  <- paste0("beta",i-1); assign(nam, 0.1)} # initial beta values
  # note the start time of the algorithm
  EM_start_time <- Sys.time()
  # run algorithm iteratively here until convergence i.e. absolute difference between iterations < 0.01
  for(k in 1:1000){
  sigmaOld <- sigma 
  for(j in 1:8){
    nam1 <- paste0("beta",j-1,"old"); nam2 <- paste0("beta",j-1)
    assign(nam1,get(nam2))
  }
    ## update
    f_value <- EM_func(input_data = data_pew_b,
                       sigma_ini=sigma,
                       beta_ini=c(beta0,beta1,beta2,
                                  beta3,beta4,beta5,
                                  beta6,beta7))
    # get sigma
    sigma <- round(f_value$sigma_hat,4)
    # get beta
    b <- round(f_value$beta_hat,4)
    # assign updated values
    for(j in 1:8){
      nam  <- paste0("beta",j-1); assign(nam, as.numeric(b[j]))
    }
    # condition for convergence
    if(abs(sigma - sigmaOld) < 0.01 & abs(beta0 - beta0old) < 0.01 & 
       abs(beta1 - beta1old) < 0.01 & abs(beta2 - beta2old) < 0.01 &
       abs(beta3 - beta3old) < 0.01 & abs(beta4 - beta4old) < 0.01 &
       abs(beta5 - beta5old) < 0.01 & abs(beta6 - beta6old) < 0.01 &
       abs(beta7 - beta7old) < 0.01) {
        EM_end_time <- Sys.time()
        print(EM_end_time - EM_start_time)
        break
    }
  }
  # function ends here
  # print final values 
  print(paste0("EM algorithm total iterations:",k))
  print(paste0("EM estimates: ","Sigma: ",sigma,
               ", Beta0: ",beta0,", Beta1: ",beta1,
               ", Beta2: ",beta2,", Beta3: ",beta3,
               ", Beta4: ",beta4,", Beta5: ",beta5,
               ", Beta6: ",beta6,", Beta7: ",beta7)) 

  # calculate x'beta with new EM parameter estimates
  x_beta_EM <- x_beta_func(data_name = cps, 
                           bhat0=beta0, bhat1=beta1, bhat2=beta2, bhat3=beta3, bhat4=beta4, bhat5=beta5, bhat6=beta6, bhat7=beta7)
  # calculate probability
  cps$yij_ebp_b <- exp(x_beta_EM)/(1+exp(x_beta_EM))
  
  ## Step 4: function to calculate weighted estimate at area level
    area_level_stat <- function(state_name,data_name)
  {
    # subset data_b by state
    data_state <- filter(data_name,state==state_name)
    # calculate weighted sum
    state_est <- round(100*sum(data_state$yij_ebp_b*data_state$weight) / sum(data_state$weight),4)
    all_state_est <- cbind(as.character(state_name),state_est)
    res <- rbind(data.frame(), all_state_est)
    return(res)
  }
  
  ## Step 5: apply above function on CPS to get EBP Yi bar using new EM parameter estimates
  
  stat_est_b <- bind_rows(lapply(cps_state_list, area_level_stat, data_name=cps))
  
  # This was the estimated value for MSPE. Next we do true value i.e. assuming model is true.

  ## Step 6: generate CPS bootstrap sample using function as before
 
  data_cps_b <- bind_rows(lapply(cps_state_list, state_func, data_name=cps))
  print(sum(data_cps_b$demvote))
  data_cps_b <- data_cps_b %>% rename("yij_ebp_b"="demvote")

  ## Step 7: apply function on CPS to get EBP Yi bar
  
  stat_true_b <- bind_rows(lapply(cps_state_list, area_level_stat, data_name = data_cps_b))
  
  # This was the true value i.e. assuming model is true.
  
  boot_out <- list()
  # Store true and estimated value for MSPE calculation
  boot_out$stat_est_b <- stat_est_b
  boot_out$stat_true_b <- stat_true_b
  
  # Store EM estimates
  boot_out$EM_est <- data.frame(sigma = sigma, Beta0 = beta0, Beta1 = beta1, Beta2 = beta2, Beta3 = beta3, Beta4 = beta4, Beta5 = beta5, Beta6 = beta6, Beta7 = beta7)
  
  # Store GLMER estimates
  logis_est <- round(fixef(logis_fit_boot),4)
  boot_out$GLMER_est <- data.frame(sigma = round(attributes(VarCorr(logis_fit_boot)$state)$stddev,4),
  Beta0 = logis_est[1], Beta1 = logis_est[2], Beta2 = logis_est[3], Beta3 = logis_est[4], Beta4 = logis_est[5], Beta5 = logis_est[6], Beta6 = logis_est[7], Beta7 = logis_est[8])
  
  return(boot_out)
}

# function for bootstrap ends here

```

# Run bootstrap function assigning number of bootstraps B

```{r }

B <- 500

boot_start_time <- Sys.time()
boot_result <- lapply(1:B,boot_func)
boot_end_time <- Sys.time()
print(paste0("Bootstrap time: ",boot_end_time - boot_start_time))

```

# Calculate MSPE with formula as mean of the values (true-est)^2

```{r }
# extract objects from list
est_val <- data.frame(lapply(boot_result, '[[',1)) # estimated value
true_val <- data.frame(lapply(boot_result, '[[',2)) # true value
# convert to numeric
est_new <- est_val[,seq(2,ncol(est_val),2)] %>% mutate_if(is.character, as.numeric)
true_new <- true_val[,seq(2,ncol(true_val),2)] %>% mutate_if(is.character, as.numeric)

write_csv(cbind(est_val[,1],est_new),"boot_est_500.csv")
write_csv(cbind(true_val[,1],true_new),"boot_true_500.csv")

MSPE <- (est_new - true_new)^2
final_MSPE <- cbind(est_val[,1],rowMeans(MSPE))  

```
